Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-21 mr-review.Rmd
revised multi-SNP text

Recognition of complex traits' diverse genetic architectures has fueled development of two-sample MR 
methods with multiple SNP instruments [@hormozdiari_leveraging_2018;@watanabe_global_2019;@pierce_power_2011;@bowden_mendelian_2015]. 
Choosing a subset of the available SNPs to serve as instruments is one issue that arises 
in MR studies. The solution to this task differs among published methods. Early methods chose a 
set of independent SNPs [@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-2;@zeng_causal_2019-1]. However, this approach can be problematic when there are multiple causal 
SNPs in LD with each other. In this setting, discarding some of the causal SNPs may only capture a small
proportion of trait variance explained by the exposure and lead to a power loss in MR [@burgess_review_2017;@burgess_mendelian_2013;@burgess_combining_2016;@burgess_bias_2011]. In support of
this is the finding that including multiple correlated SNPs in transcriptome-wide association
studies improves power compared to using independent SNPs only [@liu2021multi;@gusev_integrative_2016;@yuan_testing_2020;@zeng_non-parametric_2017].

@@

In the case of continuous exposures and outcomes, researchers have taken several approaches to extend the methods and models for 
single-SNP MR to accommodate multiple SNPs. For two stage analyses, one may perform 
a sequence of two linear regressions, as in the single-SNP MR models (Equations \@ref(eq:lm3) and \@ref(eq:lm4)). 
For an analysis
with $S$ SNP instruments, one performs two stage analysis by first regressing the 
exposure $n$ vector $X$ on the $n$ by $S + 1$ matrix of SNP instruments (and an intercept). Taking the fitted exposure values 
from this regression, the second stage involves regressing the outcome variable $Y$ on 
the fitted values, $\hat X$.

\begin{equation}
  X  = Z\alpha + \epsilon
  (\#eq:lm3)
\end{equation}

where $\epsilon$ is assumed to be a vector of independent and identically distributed normal random variables with mean zero.

\begin{equation}
  Y = \hat X\beta + \tau
  (\#eq:lm4)
\end{equation}

and $\tau$ is a vector of independent and identically distributed normal random errors with mean zero. Note also that $\tau$ in Equation \@ref(eq:lm4) is independent of $\epsilon$ in Equation \@ref(eq:lm3).

>>>>

Recognition of complex traits' diverse genetic architectures has fueled development of two-sample MR 
methods with multiple SNP instruments [@hormozdiari_leveraging_2018;@watanabe_global_2019;@pierce_power_2011;@bowden_mendelian_2015]. 
Choosing a subset of the available SNPs to serve as instruments is one issue that arises 
in MR studies. The solution to this task differs among published methods. Early methods chose a 
set of independent SNPs [@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-2;@zeng_causal_2019-1]. However, this approach can be problematic when there are multiple causal 
SNPs in linkage disequilibrium with each other. In this setting, discarding some of the causal SNPs may only capture a small
proportion of trait variance explained by the exposure and lead to a power loss in MR [@burgess_review_2017;@burgess_mendelian_2013;@burgess_combining_2016;@burgess_bias_2011]. In support of
this is the finding that including multiple correlated SNPs in transcriptome-wide association
studies improves power compared to using independent SNPs only [@liu2021multi;@gusev_integrative_2016;@yuan_testing_2020;@zeng_non-parametric_2017].

@@

@smith_mendelian_2004 recognized that many genetic variants individually explain a 
small proportion of the variation in a trait, and, thus, sufficient statistical 
power would require sample sizes in the tens of thousands 
[@schatzkin2009mendelian;@burgess_mendelian_2013]. @palmer_using_2012 partially 
resolved this issue by proposing use of multiple SNP instruments in a single MR 
analysis. @pierce_power_2011 reasoned that a single causal estimate derived from 
a collection of causal SNPs would have greater precision than an estimate derived 
from only one SNP [@burgess_mendelian_2013]. At the same time, sharing of GWAS 
summary statistics was encouraged as a best practice [@nature2012asking]. 
@harbord_severity_2013 recognized that one could calculate the ratio estimate of 
the causal effect from GWAS summary statistics, and @burgess_mendelian_2013 devised 
a strategy for determining a single causal effect from summary statistics for a 
collection of SNPs. @burgess_mendelian_2013 termed their method "inverse variance 
weighting" because each causal SNP's effect is weighted by the inverse of the variance 
of its marginal (univariate) causal effect, and the overall causal effect is the sum of the weighted SNP
causal effects.
